---
title: "Task1"
author: "Elena Chandogova"
date: "20/10/2025"
format:
  html:
    embed-resources: true
editor: source
---

## Question 1 Atomic types

Question description

```{r}
# code for solving task
# numeric
population <- 40000
area <- 60
avg_temp <- 8
bridges <- 1

#integer
schools <- 2
parks <- 3
churches <- 2

#logical
green_buildings <- FALSE
UNESCO <- FALSE

#character
city_name <- "Madion"
river_name <- "Sin"
famous_monument <- "Museum"
```


## Question 2 Vectors

Question description

```{r}
# code for solving task
#numeric vector
num_vector <- c(population, area, avg_temp, bridges, schools, parks, bridges)

#sequence vector
s_vector <- seq(from = 0, to = 100, by = 5)

#logical vector
l_vector <- sample(c(TRUE,FALSE),size=10, replace = TRUE)

```
## Question 3 Matrix

Question description

```{r}
# code for solving task
x <- sample(-50:50, 9, replace = TRUE)
m <- matrix(x, nrow = 3, ncol = 3)
rownames(m) <- c("R1","R2","R3")
colnames(m) <- c("C1","C2","C3")

row_means <- rowMeans(m)
col_means <- colMeans(m)
```
## Question 4 List

Question description

```{r}
# code for solving task
T_list <- list(population = population,
                area=area,
               avg_temp=avg_temp,
               parks=parks,
               bridges=bridges,
               schools=schools,
               UNESCO=UNESCO,
               green_buildings=green_buildings,
               city_name=city_name,
               river_name=river_name,
               num_vector=num_vector<-c(population, area, avg_temp, bridges, schools, parks, bridges),
               s_vector=s_vector <- seq(from = 0, to = 100, by =5),
               l_vector=l_vector <- sample(c(TRUE,FALSE),size=10, replace = TRUE),
               m=m,
               T_descript = 'Madion, mesto fantazie')

```
## Question 5 Factor

Question description

```{r}
# code for solving task
development_level <- factor(c('very low','low', 'medium','high','very high'), levels = c('very low','low', 'medium','high','very high'))

table(development_level)

```
## Question 6 Data frame

Question description

```{r}
# code for solving task
cities_data <- data.frame(
  cities = c('Madien', 'Banska Bystrica', 'Topolcany', 'Presov', 'Nitra', 'Bratislava'),
  population = c(40000, 73040, 27124, 91638, 78000, 475000),
  area = c(60, 103, 27, 70, 100, 367),
  unesco = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)
)

colnames(cities_data) <- c('City', 'Population', 'Area', 'UNESCO')

str(cities_data)
summary(cities_data)

cities_data$Density <- cities_data$Population / cities_data$Area

cities_data
```

##TASK 2: Subsetting & Missing Data
"Elena Chandogova"
"10/11/2025"
### 2.1. Querying Vectors

```{r}
# code for solving task
# 2.1.1. Sequence vector
# 3th element
third_el <- s_vector[3]
third_el

# 2 to 5
two_five <- s_vector[2:5]
two_five

#greater than 20
greater_20 <- s_vector[s_vector > 20]
greater_20

# 2.1.2. Logical vector
# true values
true_vl <- l_vector[l_vector]
true_vl

# index
true_pst <- which(l_vector)
true_pst
```

### 2.2 Querying Your "Madion" List
```{r}
# code for solving task
# Using the '$' operator
T_list$T_descript

# matrix [[]]
T_list[['m']]

# second element
T_list[["num_vector"]][2]

```

### 2.3. Investigating Your Data Frame
```{r}
# code for solving task
# first 3 rows
head(cities_data, 3)

# 4th row, 1st column
cities_data[4, 1]

# contains logical value
cities_data$UNESCO
```

### 2.4. Handling "Corrupted" Data (`NA`)
```{r}
# code for solving task
#copying data
df_corrupted <- cities_data
#NA
df_corrupted[2, "Area"] <- NA
df_corrupted[5, "Population"] <- NA
df_corrupted[6, "UNESCO"] <- NA
#count NA
sum(is.na(df_corrupted))
#count column NA
colSums(is.na(df_corrupted))
#mean of population with NA
mean(df_corrupted$Population)
#mean of population without NA
mean(df_corrupted$Population, na.rm = TRUE)
#remove NA
df_clean <- na.omit(df_corrupted)
df_clean
```

### 2.5. Bonus Challenge which
```{r}
# code for solving task
#highest population
highest_pop_index <- which.max(cities_data$Population)
cities_data[highest_pop_index, ]
#lowest destiny
lowest_density_index <- which.min(cities_data$Density)
cities_data[lowest_density_index, ]
```
## TASK 3: Expanding Your Universe

### 3.1. The Quest: Find Your Data
```{r}
# code for solving task
# import from: 'https://www.kaggle.com/datasets/juanmah/world-cities'
library(readr)
worldcities <- read_csv('worldcities.csv')
View(worldcities)

head(worldcities)
str(worldcities)
summary(worldcities)
```

### 3.2. The Import: Load Your Data
```{r}
# code for solving task
# load libraries
library(dplyr)
library(readr)

 # read data
# Load your data (replace with your file's name)
my_real_data <- read_csv("worldcities.csv")

# first inspection
glimpse(worldcities)
head(worldcities)
summary(worldcities)
```

### 3.3 The interogation: Explore with 'dplyr'

# Question 1: Group by UNESCO status and summarise the average population
```{r}
# code for solving task
cities_data %>%
group_by(UNESCO) %>%
summarise(avg_population = mean(Population, na.rm = TRUE),
max_population = max(Population, na.rm = TRUE),
n_cities = n())
```

# Question 2: Select only the columns 'city' and 'population' from the dataset
```{r}
# code for solving task
cities_data %>%
select(City, Population)
```
# Question 3: Filter cities with a population greater than 100000
```{r}
# code for solving task
cities_data %>%
filter(Population > 10000)
```

# Question 4: Arrange the cities by area in ascending order
```{r}
# code for solving task
cities_data %>%
arrange(Area)
```

# Question 5: Create a new column 'Population_million' that shows population in millions
```{r}
# code for solving task
cities_data %>%
mutate(Population_million = Population / 1e6)
```
# Task 4. Building a “Toolkit” for Your Universe (Functions)
# 4.1. The “Greeting” Function (A Simple Start)
```{r}
greet_universe_pretty <- function(universe_name, your_name, num_cities, avg_pop) {
cat("\033[1;34m") # Bold blue text
cat("========================================\n")
cat("✨ Welcome to the Universe:", universe_name, "✨\n")
cat("----------------------------------------\n")
cat("\033[0;32m") # Green text
cat("Report prepared by:", your_name, "\n")
cat("Number of cities in dataset:", num_cities, "\n")
cat("Average city population:", round(avg_pop, 0), "\n")
cat("\033[1;34m") # Bold blue again
cat("========================================\n")
cat("\033[0m") # Reset color
}
```
# Call the function with actual data from cities_data
```{r}
greet_universe_pretty(
universe_name = "Madion",
your_name = "Elena Chandogova",
num_cities = nrow(cities_data),
avg_pop = mean(cities_data$Population)
)
```
# The “Loading” Function (A Good Practice)
```{r}
# Enhanced function to load and inspect dataset
load_and_inspect_pretty <- function(file_path, universe_name, your_name) {
  data <- read_csv(file_path)
  
# Greeting
  cat("\033[1;34m")
  cat("============================================\n")
  cat("🌟 Welcome to the Universe:", universe_name, "🌟\n")
  cat("--------------------------------------------\n")
  cat("\033[0;32m")
  cat("Report prepared by:", your_name, "\n")
  cat("Dataset loaded from:", file_path, "\n")
  cat("Number of rows:", nrow(data), "\n")
  cat("Number of columns:", ncol(data), "\n")
  cat("\033[1;34m")
  cat("============================================\n\n")
  
# Inspect dataset
  cat("--- Structure of the dataset ---\n")
  print(glimpse(data))
  
  cat("\n--- First 6 Rows ---\n")
  print(head(data))
  
  cat("\n--- Summary ---\n")
  print(summary(data))
  
  cat("\033[0m") # Reset color
  
  return(data)
}

# Call the function and save output
my_data <- load_and_inspect_pretty(
  file_path = "worldcities.csv",
  universe_name = "Madion",
  your_name = "Elena Chandogova"
)
```
# The “Query” Function (The Core of the Task)
```{r}
filter_cities_by_population <- function(data, pop_threshold) {
  result <- data %>%
    filter(Population > pop_threshold) %>%
    arrange(desc(Population)) # optional: sort from largest to smallest
  return(result)
}
# Call the function using 'cities_data' from Task 3
large_cities <- filter_cities_by_population(cities_data, pop_threshold = 80000)
# Print the results
print(large_cities)
```
# The “Reporting” Function
```{r}
generate_report <- function(data, city_name) {
  
  # 1. Filter for the selected city
  city_data <- data %>%
    filter(City == city_name)
  
  # Check if city exists
  if(nrow(city_data) == 0){
    cat("City", city_name, "not found in the dataset.\n")
    return(NULL)
  }
  
  # 2. Calculate key statistics
  pop <- city_data$Population
  area <- city_data$Area
  density <- city_data$Density
  unesco <- city_data$UNESCO
  record_count <- nrow(city_data)
  
  # 3. Print the report
  cat("========================================\n")
  cat("📊 Report for:", city_name, "\n")
  cat("----------------------------------------\n")
  cat("Number of Records:", record_count, "\n")
  cat("Population:", pop, "\n")
  cat("Area (km²):", area, "\n")
  cat("Population Density:", round(density, 2), "\n")
  cat("UNESCO Site:", ifelse(unesco, "Yes", "No"), "\n")
  cat("========================================\n\n")
}

# Call the function for two cities
generate_report(cities_data, "Bratislava")
generate_report(cities_data, "Nitra")
```
# Installing
```{r}
# 2. Načítanie balíka
library(ggplot2)
```
# Relationship Map: Scatter Plot of Population vs Area
```{r}
ggplot(data = cities_data, aes(x = Area, y = Population)) +
  geom_point(alpha = 0.7, size = 3, colour = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "gray") +
  labs(title = "Population vs Area of Cities",
       subtitle = "Relationship between city size and population",
       x = "Area (km²)",
       y = "Population") +
  theme_minimal()
```
# Comparison: Boxplot of Population by UNESCO status
```{r}
ggplot(data = cities_data, aes(x = UNESCO, y = Population, fill = UNESCO)) +
  geom_boxplot() +
  labs(title = "Comparison of Population by UNESCO Status",
       x = "UNESCO Site",
       y = "Population") +
  theme_light() +
  theme(legend.position = "none")
```
# Deep Dive: Faceted Histogram of Population by UNESCO status
```{r}
ggplot(data = cities_data, aes(x = Population)) +
  geom_histogram(bins = 10, fill = "coral", color = "black") +
  facet_wrap(~ UNESCO) +
  labs(title = "Distribution of Population by UNESCO Status",
       x = "Population",
       y = "Count") +
  theme_minimal()
```
# Exporting the favourite plot
```{r}
scatter_plot <- ggplot(data = cities_data, aes(x = Area, y = Population)) +
  geom_point(alpha = 0.7, size = 3, colour = "blue", na.rm = TRUE) +
  geom_smooth(method = "loess", se = TRUE, color = "gray", na.rm = TRUE) +
  labs(title = "Population vs Area of Cities",
       subtitle = "Relationship between city size and population",
       x = "Area (km²)",
       y = "Population") +
  theme_minimal()

scatter_plot
```